Optimized Keyword Persistent Hash Maps for Clojure

### 2016-12-06 paper/paper.tex
paper

Hash Array Mapped Tries (HAMT) have rocked the functional programming
world with a fast, immutable and persistent alternative
to a hash map.
They are featured in mainstream functional programming
languages like Clojure and Scala, and have been ported
to many others.

HAMT's compare very well to other similar data structures.
They offer fast lookups by minimising the depth and
branching factors of their tree representation.
They minimize memory usage by lazily creating subtries
only when necessary.

@@

A \textit{trie} is a way of formatting key/value pairs
in a tree, where values are leaves and keys are spread
across the paths to those nodes.
Key prefixes occur on the shallow levels of the tree,
and suffixes occur closer to the leaves.

A \textit{bit trie} assumes the mapping keys are strings
of bits. Each level consumes one or more bits to index
its elements.

An \textit{Array Mapped Trie}

In this paper, we explore Clojure's persistent hash
map implementation.
It was originally implemented by Hickey, based on
Bagwell's original formulation~\cite{bagwell2001ideal},
extended to be persistent.
Persistent data structures use \textit{structural sharing}
when extending themselves, so Clojure necessarily enforces
hash maps to be \textit{immutable}.

@@

\begin{enumerate}
  \item We summarize the history of Hash Array Mapped Tries (HAMT).
  \item We give an approachable walkthrough of mechanics behinds
      HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation,
    via our pure-Clojure reimplementation.
  \item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  \item We speculate on future directions for HAMT's.
\end{enumerate}

\section{History of Hash Mapped Array Tries}

First described by Bagwell~\cite{bagwell2001ideal},
Hickey showed how to make a \textit{persistent}

HMap via structural sharing, further reducing
memory usage.

\section{Walkthrough}

@@

\section{Characteristics of HAMT's}

\section{Clojure's HAMT, in detail}

@@

@@

\subsection{Deletion}

@@

We reimplemented the \texttt{PersistentHashMap} class from
the Clojure standard library in Clojure---it was originally
written in Java.
This is not the first reimplementation. ClojureScript
features similar pure-Clojure port in its standard library,
and several other languages have ported their own versions
based on Clojure's original implementation.

@@

\section{Experiments}

@@

\begin{figure}
\begin{verbatim}
(assoc [this key val]
  (cond
    (keyword? key)
    (case key
      ...   ;; specific branches for each key
      ;; default case
      )

    :else
\end{verbatim}
\caption{Psuedocode for \texttt{assoc} in optimized maps}
\end{figure}

\begin{figure}
\begin{verbatim}
(get [this key]
\end{verbatim}
\caption{Psuedocode for \texttt{get} in optimized maps}
\end{figure}

\begin{figure*}
\includegraphics[width=18cm]{big-tree}
\caption{
Trie visualization in Clojure port, for map
\texttt{\{:a 1 :b 2 :c 3 :d 4 :e 5 :f 6 :g 7 :h 8 :i 9 :j 10\}}
(See Appendix~\ref{hash-examples} for corresponding hashes).
}
\end{figure*}

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
      legend pos=north west,
      xlabel=Number of Extra keys,
      ylabel=Time (msec)]
      \addplot+[sharp plot] coordinates
      {(0,837.900884) (100,1435.768863) (200,1447.167303) (300,1505.881845) (400,1331.124945) 
        (500,1759.42348 8) (600,1387.997231) (700,1729.756068) (800,1784.948714) (900,1481.740526) 
        (1000,1521.329677) (1100,1855.281003) (1200,2573.8774) (1300,2049.42444) (1400,1817.21074) 
        (1500,2023.47916) (1600,1936.875666) (1700,2129.781288) (1800,1757.819692) (1900,2973.424349)};
      \addlegendentry{Plain map}
      \addplot+[sharp plot] coordinates
{
(0,840.862397) (100,894.77844) (200,921.830137) (300,904.195596) (400,868.788413) 
(500,849.688625) (600,857.498675) (700,884.393927) (800,1099.811693) (900,936.40417) 
(1000,1079.314484) (1100,1598.098832) (1200,1334.392182) (1300,1212.918047) (1400,1047.268085) 
(1500,1110.733413) (1600,1232.646281) (1700,1181.103518) (1800,1400.642201) (1900,2108.923538)
};
      \addlegendentry{Record}
      \addplot+[sharp plot] coordinates
{
(0,892.363186) (100,972.581187) (200,921.286823) (300,944.139623) (400,850.684909) 
(500,907.794038) (600,835.013332) (700,841.138563) (800,938.415172) (900,960.614487)
 (1000,904.91121) (1100,1569.635548) (1200,1294.22056) (1300,1047.200465) (1400,1112.890638)
 (1500,1203.444674) (1600,1261.062708) (1700,2109.647289) (1800,3016.390694) (1900,3844.42986)
};
      \addlegendentry{Optimized map}
    \end{axis}
  \end{tikzpicture}
  \caption{Running time for keyword keys benchmark.}
\end{figure}

@@

We tested against

\begin{figure}
\begin{verbatim}
(defn exercise-bench [f n]
  (loop [i 20
         m (f (into {:a 1 :b 2 :c 3 :d 4
                     :e 5 :f 6 :g 7 :h 8}
                    (map #(vector % %) (range n))))]
    (when-not (zero? i)
      (dotimes [_ 100000] 
        (+ (:a m) (:b m) (:c m) (:d m)
           (:e m) (:f m) (:g m) (:h m)))
      (recur (dec i) (update m :a inc)))))
\end{verbatim}
\caption{Benchmark code that takes a function $f$
that returns the kind of map we are currently
benchmarking (eg., plain, record, or optimized).
}
\end{figure}

@@

@@

>>>>

Hash Array Mapped Tries (HAMT) have rocked the functional programming
world with a fast, immutable and persistent alternative
to a hash map.
First described by Bagwell~\cite{bagwell2001ideal},
they are featured in mainstream functional programming
languages like Clojure and Scala, and have been ported
to many others.

HAMT's compare very well to other similar data structures.
They offer fast lookups by minimizing the depth and
branching factors of their tree representation.
They minimize memory usage by lazily creating subtries
only when necessary.

@@

To understand with a hash array mapped trie is, 
we first give some definitions.
A \textit{trie} is a way of formatting key/value pairs
in a tree, where values are leaves and keys are spread
across the paths to those nodes.
Key prefixes occur on the shallow levels of the tree,
and suffixes occur closer to the leaves.
A \textit{bit trie} assumes the mapping keys are strings
of bits. Each level consumes one or more bits to index
its elements.
An \textit{Array Mapped Trie}
maps the bits of array indices as a bit trie.

In this paper, we explore Clojure's persistent hash
map implementation
(Section~\ref{clojure-hamt}).
It was implemented by Hickey~\cite{hickey2008clojure}, extending
Bagwell's original formulation~\cite{bagwell2001ideal}
to be persistent.
Persistent data structures use \textit{structural sharing}
when extending themselves, so Clojure necessarily enforces
hash maps to be \textit{immutable}.

@@

\begin{enumerate}
  \item We give an approachable walkthrough of mechanics behinds
      HAMTs.
  \item We describe the internals of Clojure's persistent HAMT implementation,
    via our pure-Clojure reimplementation.
  \item We present a port of Clojure's HAMT from Java to Clojure
    for pedagogical purposes,
    featuring trie visualization (Figure~\label{port-visualize}).
  \item We evaluate a Clojure-specific optimization for Clojure's HAMT.
  \item We speculate on future directions for HAMT's.
\end{enumerate}

\section{Walkthrough}
\label{walkthrough}

@@

%\section{Characteristics of HAMT's}

\section{Clojure's HAMT, in detail}
\label{clojure-hamt}

@@

Search is straightforward, and works as described
in Section~\ref{walkthrough}.

@@

\subsection{Deletion}
Deletion is slightly more complicated than Bagwell's original
HAMT. Since Clojure's HAMT is persistent and immutable, it
must copy and alter the node containing the deleting leaf.
The rest of the trie can be reused, making this still a cheap
operation.

@@

We reimplemented~\footnote{\url{https://github.com/frenchy64/optimized-kw-maps}}
the \texttt{PersistentHashMap} class from
the Clojure standard library in Clojure---it was originally
written in Java.
This is not the first reimplementation. ClojureScript
features similar pure-Clojure port in its standard library,
and several other languages have ported their own versions
based on Clojure's original implementation.

@@

\subsection{Visualization}

We also implemented graphical trie visualization 
via \texttt{rhizome}
\footnote{\url{https://github.com/ztellman/rhizome}}.
See Figure~\ref{big-trie-vis} for an example
for a representative trie.

\section{Experiments}
\label{experiments}

@@

%\begin{figure}
%\begin{verbatim}
%(assoc [this key val]
%  (cond
%    (keyword? key)
%    (case key
%      ...   ;; specific branches for each key
%      ;; default case
%      )
%
%    :else
%\end{verbatim}
%\caption{Pseudocode for \texttt{assoc} in optimized maps}
%\end{figure}
%
%\begin{figure}
%\begin{verbatim}
%(get [this key]
%  (cond
%    (keyword? key)
%    (case key
%      ...   ;; specific branches for each key
%      ;; default case
%      )
%
%    :else
%\end{verbatim}
%\caption{Pseudocode for \texttt{get} in optimized maps}
%\end{figure}

\begin{figure*}
\includegraphics[width=18cm]{big-tree}
\caption{
Trie visualization in Clojure port, for map
\texttt{\{:a 1 :b 2 :c 3 :d 4 :e 5 :f 6 :g 7 :h 8 :i 9 :j 10\}}
(See Appendix~\ref{hash-examples} for corresponding hashes).
}
\label{big-trie-vis}
\end{figure*}

\begin{figure}
  \begin{tikzpicture}
    \begin{axis}[
      legend pos=north west,
      xlabel=Number of Extra keys,
      ylabel=Time (msec)]
      \addplot+[sharp plot] coordinates
      {(0,837.900884) (100,1435.768863) (200,1447.167303) (300,1505.881845) (400,1331.124945) 
        (500,1759.42348 8) (600,1387.997231) (700,1729.756068) (800,1784.948714) (900,1481.740526) 
        (1000,1521.329677) (1100,1855.281003) (1200,2573.8774) (1300,2049.42444) (1400,1817.21074) 
        (1500,2023.47916) (1600,1936.875666) (1700,2129.781288) (1800,1757.819692) (1900,2973.424349)};
      \addlegendentry{Plain map}
      \addplot+[sharp plot] coordinates
{
(0,840.862397) (100,894.77844) (200,921.830137) (300,904.195596) (400,868.788413) 
(500,849.688625) (600,857.498675) (700,884.393927) (800,1099.811693) (900,936.40417) 
(1000,1079.314484) (1100,1598.098832) (1200,1334.392182) (1300,1212.918047) (1400,1047.268085) 
(1500,1110.733413) (1600,1232.646281) (1700,1181.103518) (1800,1400.642201) (1900,2108.923538)
};
      \addlegendentry{Record}
      \addplot+[sharp plot] coordinates
{
(0,892.363186) (100,972.581187) (200,921.286823) (300,944.139623) (400,850.684909) 
(500,907.794038) (600,835.013332) (700,841.138563) (800,938.415172) (900,960.614487)
 (1000,904.91121) (1100,1569.635548) (1200,1294.22056) (1300,1047.200465) (1400,1112.890638)
 (1500,1203.444674) (1600,1261.062708) (1700,2109.647289) (1800,3016.390694) (1900,3844.42986)
};
      \addlegendentry{Optimized map}
    \end{axis}
  \end{tikzpicture}
  \caption{Running time for keyword keys benchmark,
  varying the number of extra entries in the map.}
  \label{bench-plot}
\end{figure}

@@

We tested three kinds of maps in our benchmark.

\paragraph{Plain map}
This is the Clojure implementation of hash array mapped trie
described in Section~\ref{clojure-hamt}.

\paragraph{Record}
This is a map constructed with Clojure's \texttt{defrecord}
form. It is specialized at compile time for a specific
known set of keyword entries.

\paragraph{Optimized map}
This is our prototype implementation that specializes
maps at runtime.

\begin{figure}
\begin{verbatim}
(defn exercise-bench [f n]
  (loop [i 20
         m (f (into {:a 1 :b 2 :c 3 :d 4
                     :e 5 :f 6 :g 7 :h 8}
                    (map #(vector % %) (range n))))]
    (when-not (zero? i)
      (dotimes [_ 100000] 
        (+ (:a m) (:b m) (:c m) (:d m)
           (:e m) (:f m) (:g m) (:h m)))
      (recur (dec i) (update m :a inc)))))
\end{verbatim}
\caption{Benchmark code that takes a function $f$
that returns the kind of map we are currently
benchmarking (eg., plain, record, or optimized),
and a number $n$, the number of extra keys to fill
the map with.
}
\label{benchmark-code}
\end{figure}

Figure~\ref{benchmark-code} contains the benchmark
code. The purpose of the benchmark is to simulate
a set of map associations followed by many lookups,
since our optimization only triggers after an assocation
operation.
In this version of the benchmark, we perform 20
associations, each with 400,000 lookups
on keyword entries.
Each iteration has a set number of keyword entries,
and we vary the

\subsection{Results}

Figure~\ref{bench-plot} plots the results.
The \emph{Record} is hand-optimized for the particular
keyword entries in the benchmark, and outperforms
the \emph{Plain map} as expected.
Our \emph{Optimized map} compiles its own specialized
map for the set of keywords, and consistently
beats \emph{Plain map} for 1,500 extra keys and under.
It also stays competitive with the hand-tuned \emph{Record}
for 1,500 and under.

After 1,500 keys, there is an unknown problem with our optimized
implementation. Our hand-written benchmarking scheme
does not compensate for garbage collection and is rather crude,
so we would like to revisit this benchmark in the future to 
investiate this issue.

\subsection{Interpretation}

Our optimization stays mostly competitive with hand-tuned
records. However, there are no wins in our predicted
primary use-case: small maps with only keyword keys.
Our benchmark tested an unusual case: keyword lookups on a
hybrid keyword and integer keyed map.
We constructed it like this to ensure our implementation beat
plain maps in expected cases.

We hope our approach can be improved to cater to our
original vision of speeding up small keyword maps---our
initial experiments show no speedup in this case.

@@

In future work, we plan to devise several benchmarks
that test real-world usages of small keyword maps.
We will devise several more optimization strategies
based on the optimization described in Section~\ref{experiments},
and evaluate them using these new benchmarks.

Furthermore, we plan to override Clojure's default hash map implementation
in a custom version of Clojure, and run benchmarks
over entire programs.
We plan to find programs that extensively use the ``tagged map''
idiom of using plain maps as records---we conjecture these
are more likely to see speedups if our optimizations
satisfy the micro-benchmarks.

@@

Clojure has a general purpose hash-map implementation, but
programmers are encouraged to hash-maps primarily as records,
which are typically small maps with less than 32 entries
with only keyword keys.

In this paper, we prototyped an optimization that caters
to this usecase, by compiling specialized hash-maps
for specific keyword keysets discovered at runtime.
Our preliminary investigation found we could duplicate
performance for hand-tuned records with hybrid keyword and
integer keyed maps.
In the future, we plan to investigate optimizations
and benchmarks that better target small keyword maps.
